---
title: "Total B cell abundance exhibits stronger predictive power than subtypes"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
require(ggplot2)
ggplot2::theme_set(new = theme_bw(base_size = 12)) # sets the theme and axis font size
```


# Aims:

- Study the additional questions (2): Could specific B cell subsets offer stronger predictive power compared to overall B cell abundance?


# load required package
```{r, include=FALSE}
library(Seurat)
library(dplyr)
library(ggplot2)


library(pROC)
library(verification)

library(purrr)
library(readr)
library(readxl)
library(survival)
library(survminer)
library(UCell)

```

# set color
```{r}

custom_colors = c("#1c4f27", "#81144e", "#79831e", "#00305d", "#C0C0C0", "#9c1915", "black", "#404040", "#808080", "#D3D3D3")

```

# set input and output directories & parameters
```{r}

result_fig_dir = "../03.Results/Figures/"

```

# common functions and constants
```{r}

signatures <- list(
    Switched_UCell_sig = c("IGHA1", "IGHA2", "IGHG1","IGHG2","IGHG3","IGHG4"), 
    Unswitched_UCell_sig = c("IGHD","IGHM")
)

```

# Cohort6 test class-switched vs non-class-switched B cells on ICB response
```{r}

GSE234933_seu = readRDS("../02.Input/GSE234933/seu_Tissue.rds")
GSE234933_seu <- AddModuleScore_UCell(GSE234933_seu, features=signatures, name=NULL)

colnames(GSE234933_seu@meta.data)

switched_indice = (GSE234933_seu$Switched_UCell_sig > GSE234933_seu$Unswitched_UCell_sig)
Bcell_indice = (GSE234933_seu$cellType_origin_L2 == "B_cells")

GSE234933_seu$Switched_B = (switched_indice & Bcell_indice)
GSE234933_seu$Unswitched_B = (!(switched_indice) & Bcell_indice)

# calculate ratios of Switched_B and Unswitched_B in each patient
meta_data <- GSE234933_seu@meta.data
ratios <- meta_data %>%
  group_by(patient) %>%
  summarise(
    total_cells = n(),
    B_cells = sum(cellType_origin_L2 == "B_cells"),
    switched_cells = sum(Switched_B == 1),
    unswitched_cells = sum(Unswitched_B == 1),
    NaiveB_cells = sum(cellType_origin_L3 == "NaiveB"),
    MemoryB_cells = sum(cellType_origin_L3 == "MemoryB"),
    PlasmaB_cells = sum(cellType_origin_L3 == "PlasmaB"),
    B_cells_ratio = B_cells / total_cells,
    switched_ratio = switched_cells / total_cells,
    unswitched_ratio = unswitched_cells / total_cells,
    NaiveB_ratio = NaiveB_cells / total_cells,
    MemoryB_ratio = MemoryB_cells / total_cells,
    PlasmaB_ratio = PlasmaB_cells / total_cells
  )

write.csv(ratios, file = "../02.Input/GSE234933/B_cells_subgroup_ratios.csv")
saveRDS(GSE234933_seu, "../02.Input/GSE234933/seu_Tissue.rds")


## load B cell subgroup ratios
Bratio_df = read.csv(file = "../02.Input/GSE234933/B_cells_subgroup_ratios.csv", row.names = 1)

## load phenotype information
GSE234933_clinicalInfo_df = read.csv('../02.Input/GSE234933/GSE234933_Clinical_Info.csv', header=TRUE, row.names = 1)
GSE234933_clinicalInfo_df = na.omit(GSE234933_clinicalInfo_df)
GSE234933_clinicalInfo_df$PatientID = rownames(GSE234933_clinicalInfo_df)
GSE234933_clinicalInfo_df_new = GSE234933_clinicalInfo_df[c("PatientID","HPV.Status","Original.anatomic.site","Anatomic.location.of.scRNA.seq.specimen","Sex","Age","Smoking.history","ICB_response2")]
colnames(GSE234933_clinicalInfo_df_new) = c("PatientID","hpv_status","Tumorsite","scRNAseq_loc","Sex","Age","TobaccoUse","ICB_R")

# merge ICF info df with clinical info df
GSE234933_clinicalInfo_df = GSE234933_clinicalInfo_df_new
sampleNames = Bratio_df$patient
GSE234933_all_info <- merge(GSE234933_clinicalInfo_df, Bratio_df, by.x = "PatientID",  by.y = "patient", all.x = T)

## AUC calculation
all_info = GSE234933_all_info
cellType_all = c("B_cells_ratio", "switched_ratio", "unswitched_ratio", "NaiveB_ratio", "MemoryB_ratio", "PlasmaB_ratio")
result_df = data.frame(cellType = character(),
                       AUC = numeric(),
                       AUC_low = numeric(),
                       AUC_up = numeric(),
                       p_val = numeric())
for (cn_i in 1:length(cellType_all)){
  cn = cellType_all[cn_i]
  cancerData = all_info[c(cn, "ICB_R")]
  colnames(cancerData) = c("CellType", "ICB_R")
  logistic_model = glm(ICB_R ~ CellType, family='binomial', data=cancerData)
  predict_RecR = predict(logistic_model, cancerData['CellType'], type = "response")
  roc_result=roc(cancerData$ICB_R,cancerData$CellType, direction="<",levels=c(0, 1)) 
  auc_RecR = roc_result$auc[1]
  auc_95CI = c(ci.auc(roc_result)[1],ci.auc(roc_result)[3])
  auc_p=roc.area(cancerData$ICB_R,predict_RecR)
  p_RecR = auc_p$p.value
  result_df[cn_i,1] = cn
  result_df[cn_i,2:5] = c(auc_RecR, auc_95CI, p_RecR)
}
write.csv(result_df, "../02.Input/GSE234933/AUC_ICB_B_cell_subgroups.csv")


```

# Cohort7 (PBMC and tissue)  test class-switched vs non-class-switched B cells on ICB response
```{r}

Tissue = "PBMC" # Tissue PBMC

GSE200996_seu = readRDS(paste0("../02.Input/GSE200996/seu_",Tissue,".rds"))
GSE200996_seu <- AddModuleScore_UCell(GSE200996_seu, features=signatures, name=NULL)

colnames(GSE200996_seu@meta.data)

switched_indice = (GSE200996_seu$Switched_UCell_sig > GSE200996_seu$Unswitched_UCell_sig)
Bcell_indice = (GSE200996_seu$cell_type_L2 == "B")

GSE200996_seu$Switched_B = (switched_indice & Bcell_indice)
GSE200996_seu$Unswitched_B = (!(switched_indice) & Bcell_indice)

# calculate ratios of Switched_B and Unswitched_B in each patient
meta_data <- GSE200996_seu@meta.data
ratios <- meta_data %>%
  group_by(patient) %>%
  summarise(
    total_cells = n(),
    B_cells = sum(cell_type_L2 == "B"),
    switched_cells = sum(Switched_B == 1),
    unswitched_cells = sum(Unswitched_B == 1),
    NaiveB_cells = sum(cell_type_L3 == "NaiveB"),
    MemoryB_cells = sum(cell_type_L3 == "MemoryB"),
    PlasmaB_cells = sum(cell_type_L3 == "PlasmaB"),
    B_cells_ratio = B_cells / total_cells,
    switched_ratio = switched_cells / total_cells,
    unswitched_ratio = unswitched_cells / total_cells,
    NaiveB_ratio = NaiveB_cells / total_cells,
    MemoryB_ratio = MemoryB_cells / total_cells,
    PlasmaB_ratio = PlasmaB_cells / total_cells
  )

write.csv(ratios, file = paste0("../02.Input/GSE200996/B_cells_subgroup_ratios_",Tissue,".csv"))
saveRDS(GSE200996_seu, paste0("../02.Input/GSE200996/seu_",Tissue,".rds"))


## load B cell subgroup ratios
Bratio_df = read.csv(file = paste0("../02.Input/GSE200996/B_cells_subgroup_ratios_",Tissue,".csv"), row.names = 1)
Bratio_df$PatientID = sub("-.*", "", Bratio_df$patient)

## load phenotype information
# clinical info.
GSE200996_clinicalInfo_df = read.csv('../02.Input/GSE200996/GSE200996_PatientOutcome.csv', header=TRUE, row.names = 1)
GSE200996_clinicalInfo_df$PatientID = rownames(GSE200996_clinicalInfo_df)
GSE200996_clinicalInfo_df_new = GSE200996_clinicalInfo_df[c("PatientID","hpv_status","Tumorsite","PathologicalStage","PathologicalNode","Sex","Age","TobaccoUse","AlcoholUse")]
GSE200996_clinicalInfo_df_new$RECIST_R = GSE200996_clinicalInfo_df[,'RECIST.response.excluding.non.measurable']
GSE200996_clinicalInfo_df_new$RECIST_R[grepl('stable/progress', GSE200996_clinicalInfo_df_new$RECIST_R)] = 0
GSE200996_clinicalInfo_df_new$RECIST_R[grepl('response', GSE200996_clinicalInfo_df_new$RECIST_R)] = 1
GSE200996_clinicalInfo_df_new$RECIST_R[grepl('not measurable', GSE200996_clinicalInfo_df_new$RECIST_R)] = NA
GSE200996_clinicalInfo_df_new$RECIST_R = as.numeric(GSE200996_clinicalInfo_df_new$RECIST_R)
GSE200996_clinicalInfo_df_new$Patho_R = GSE200996_clinicalInfo_df[,'Pathological.response..'] 
GSE200996_clinicalInfo_df_new$Patho_minR = (GSE200996_clinicalInfo_df_new$Patho_R>=10)+0

# merge ICF info df with clinical info df
GSE200996_clinicalInfo_df = GSE200996_clinicalInfo_df_new
GSE200996_all_info <- merge(Bratio_df, GSE200996_clinicalInfo_df, by = "PatientID", all.x = T)

## AUC calculation
all_info = GSE200996_all_info
cellType_all = c("B_cells_ratio", "switched_ratio", "unswitched_ratio", "NaiveB_ratio", "MemoryB_ratio", "PlasmaB_ratio")
result_df = data.frame(cellType = character(),
                       AUC = numeric(),
                       AUC_low = numeric(),
                       AUC_up = numeric(),
                       p_val = numeric())
for (cn_i in 1:length(cellType_all)){
  cn = cellType_all[cn_i]
  cancerData = all_info[c(cn, "RECIST_R")] # RECIST_R Patho_minR
  colnames(cancerData) = c("CellType", "RECIST_R")
  logistic_model = glm(RECIST_R ~ CellType, family='binomial', data=cancerData)
  predict_RecR = predict(logistic_model, cancerData['CellType'], type = "response")
  roc_result=roc(cancerData$RECIST_R,cancerData$CellType, direction="<",levels=c(0, 1)) 
  auc_RecR = roc_result$auc[1]
  auc_95CI = c(ci.auc(roc_result)[1],ci.auc(roc_result)[3])
  auc_p=roc.area(cancerData$RECIST_R,predict_RecR)
  p_RecR = auc_p$p.value
  result_df[cn_i,1] = cn
  result_df[cn_i,2:5] = c(auc_RecR, auc_95CI, p_RecR)
}
write.csv(result_df, paste0("../02.Input/GSE200996/AUC_ICB_B_cell_subgroups_",Tissue,"_RECIST_R.csv"))

```

# Cohort5 (scRNAseq) test class-switched vs non-class-switched B cells on ICB response
```{r}

GSE232240_seu = readRDS("../02.Input/GSE232240/seu_Tissue.rds")
GSE232240_seu <- AddModuleScore_UCell(GSE232240_seu, features=signatures, name=NULL)

colnames(GSE232240_seu@meta.data)
GSE232240_seu$patient_time = paste(GSE232240_seu$patient, GSE232240_seu$timepoint, sep = "_")

switched_indice = (GSE232240_seu$Switched_UCell_sig > GSE232240_seu$Unswitched_UCell_sig)
Bcell_indice = (GSE232240_seu$cell_type == "B")

GSE232240_seu$Switched_B = (switched_indice & Bcell_indice)
GSE232240_seu$Unswitched_B = (!(switched_indice) & Bcell_indice)

# calculate ratios of Switched_B and Unswitched_B in each patient
meta_data <- GSE232240_seu@meta.data
ratios <- meta_data %>%
  group_by(patient_time) %>%
  summarise(
    total_cells = n(),
    B_cells = sum(cell_type_L2 == "B"),
    switched_cells = sum(Switched_B == 1),
    unswitched_cells = sum(Unswitched_B == 1),
    NaiveB_cells = sum(cell_type_L3 == "NaiveB"),
    MemoryB_cells = sum(cell_type_L3 == "MemoryB"),
    PlasmaB_cells = sum(cell_type_L3 == "PlasmaB"),
    B_cells_ratio = B_cells / total_cells,
    switched_ratio = switched_cells / total_cells,
    unswitched_ratio = unswitched_cells / total_cells,
    NaiveB_ratio = NaiveB_cells / total_cells,
    MemoryB_ratio = MemoryB_cells / total_cells,
    PlasmaB_ratio = PlasmaB_cells / total_cells
  )

write.csv(ratios, file = "../02.Input/GSE232240/B_cells_subgroup_ratios.csv")
saveRDS(GSE232240_seu, "../02.Input/GSE232240/seu_Tissue.rds")


## load B cell subgroup ratios
Bratio_df = read.csv(file = "../02.Input/GSE232240/B_cells_subgroup_ratios.csv", row.names = 1)

## load phenotype information
clinical_info = read.csv(file = paste0("../02.Input/GSE232240/clinical_info.csv"), header = T)
clinical_info$response[clinical_info$response=="NR"] = 0
clinical_info$response[clinical_info$response=="RE"] = 1
clinical_info = clinical_info[c("response", "meta")]
clinical_info$response = as.numeric(clinical_info$response)

# merge ICF info df with clinical info df
GSE232240_clinicalInfo_df = clinical_info
GSE232240_clinicalInfo_df
GSE232240_all_info <- merge(GSE232240_clinicalInfo_df, Bratio_df, by.x = "meta",  by.y = "patient_time", all.x = T)

## AUC calculation
all_info = GSE232240_all_info
cellType_all = c("B_cells_ratio", "switched_ratio", "unswitched_ratio", "NaiveB_ratio", "MemoryB_ratio", "PlasmaB_ratio")
result_df = data.frame(cellType = character(),
                       AUC = numeric(),
                       AUC_low = numeric(),
                       AUC_up = numeric(),
                       p_val = numeric())
for (cn_i in 1:length(cellType_all)){
  cn = cellType_all[cn_i]
  cancerData = all_info[c(cn, "response")]
  colnames(cancerData) = c("CellType", "response")
  logistic_model = glm(response ~ CellType, family='binomial', data=cancerData)
  predict_RecR = predict(logistic_model, cancerData['CellType'], type = "response")
  roc_result=roc(cancerData$response,cancerData$CellType, direction="<",levels=c(0, 1)) 
  auc_RecR = roc_result$auc[1]
  auc_95CI = c(ci.auc(roc_result)[1],ci.auc(roc_result)[3])
  auc_p=roc.area(cancerData$response,predict_RecR)
  p_RecR = auc_p$p.value
  result_df[cn_i,1] = cn
  result_df[cn_i,2:5] = c(auc_RecR, auc_95CI, p_RecR)
}
write.csv(result_df, "../02.Input/GSE232240/AUC_ICB_B_cell_subgroups.csv")


```

# Cohort11 test class-switched vs non-class-switched B cells on ICB response
```{r}

data_dir = "../02.Input/HNSCC-Amanda/"
Amanda_flow_all_info <- read_excel(paste0(data_dir, "HNSCC ITx annotations all assays_v3_flow_arranged_20241004.xlsx"), sheet = "Sheet1")

colnames(Amanda_flow_all_info)

Amanda_flow_all_info = Amanda_flow_all_info[c(16, 17, 21, 177:186, 606)]

Amanda_flow_all_info$ICBR = 0
Amanda_flow_all_info$ICBR[Amanda_flow_all_info$BOR %in% c("CR", "PR")] = 1 

cellType_all = c("B_cells_ratio", "switched_ratio", "unswitched_ratio", "NaiveB_ratio", "MemoryB_ratio", "PlasmaB_ratio")
Amanda_flow_all_info$B_cells_ratio = Amanda_flow_all_info$`Flow B Cells count`/(Amanda_flow_all_info$`Flow Monocytes Count` + Amanda_flow_all_info$`Flow Lymphocytes Count`)
Amanda_flow_all_info$NaiveB_ratio = Amanda_flow_all_info$`Flow B Cells Naive (CD21+CD27-) (% B Cells)`*Amanda_flow_all_info$Bratio/100
Amanda_flow_all_info$PlasmaB_ratio = Amanda_flow_all_info$`Flow Plasmablasts (CD38++IgM-) (% B Cells)`*Amanda_flow_all_info$Bratio/100
Amanda_flow_all_info$MemoryB_ratio = Amanda_flow_all_info$`Flow B Cells Memory (CD21+CD27+) (% B Cells)`*Amanda_flow_all_info$Bratio/100

switchedB = Amanda_flow_all_info$`Flow Plasmablasts (CD38++IgM-) (% B Cells)` + Amanda_flow_all_info$`Flow B Cells Switched Memory (IgM/IgD-) (% B Cells)`
unswitchedB = 100 - switchedB
Amanda_flow_all_info$switched_ratio = switchedB*Amanda_flow_all_info$Bratio/100
Amanda_flow_all_info$unswitched_ratio = unswitchedB*Amanda_flow_all_info$Bratio/100


## AUC calculation
all_info = Amanda_flow_all_info
cellType_all = c("B_cells_ratio", "switched_ratio", "unswitched_ratio", "NaiveB_ratio", "MemoryB_ratio", "PlasmaB_ratio")
result_df = data.frame(cellType = character(),
                       AUC = numeric(),
                       AUC_low = numeric(),
                       AUC_up = numeric(),
                       p_val = numeric())
for (cn_i in 1:length(cellType_all)){
  cn = cellType_all[cn_i]
  cancerData = all_info[c(cn, "ICBR")]
  colnames(cancerData) = c("CellType", "ICBR")
  logistic_model = glm(ICBR ~ CellType, family='binomial', data=cancerData)
  predict_RecR = predict(logistic_model, cancerData['CellType'], type = "response")
  roc_result=roc(cancerData$ICBR,cancerData$CellType, direction="<",levels=c(0, 1)) 
  auc_RecR = roc_result$auc[1]
  auc_95CI = c(ci.auc(roc_result)[1],ci.auc(roc_result)[3])
  auc_p=roc.area(cancerData$ICBR,predict_RecR)
  p_RecR = auc_p$p.value
  result_df[cn_i,1] = cn
  result_df[cn_i,2:5] = c(auc_RecR, auc_95CI, p_RecR)
}
write.csv(result_df, "../02.Input/HNSCC-Amanda/AUC_ICB_B_cell_subgroups.csv")


```

#  Figure S9A. AUCs boxplot comparing B cells and different B subgroups
```{r}


GSE234933_result_df = read.csv("../02.Input/GSE234933/AUC_ICB_B_cell_subgroups.csv")

GSE200996_Tissue_result_df = read.csv(paste0("../02.Input/GSE200996/AUC_ICB_B_cell_subgroups_Tissue_RECIST_R.csv"))

GSE200996_PBMC_result_df = read.csv(paste0("../02.Input/GSE200996/AUC_ICB_B_cell_subgroups_PBMC_RECIST_R.csv"))

GSE232240_result_df = read.csv("../02.Input/GSE232240/AUC_ICB_B_cell_subgroups.csv")

Amanda_result_df = read.csv("../02.Input/HNSCC-Amanda/AUC_ICB_B_cell_subgroups.csv")


merged_df <- bind_rows(
  GSE234933_result_df %>% mutate(dataset = "Cohort6"),
  GSE200996_Tissue_result_df %>% mutate(dataset = "Cohort7 (tissue)"),
  GSE200996_PBMC_result_df %>% mutate(dataset = "Cohort7 (PBMC)"),
  GSE232240_result_df %>% mutate(dataset = "Cohort5"),
  Amanda_result_df %>% mutate(dataset = "Cohort11")
)

merged_df$dataset = factor(merged_df$dataset, levels = c("Cohort5", "Cohort6", "Cohort7 (tissue)", "Cohort7 (PBMC)", "Cohort11"))


# Calculate the median AUC for each cellType and reorder the x-axis based on it
merged_df <- merged_df %>%
  group_by(cellType) %>%
  mutate(median_AUC = median(AUC)) %>%
  ungroup()

merged_df$cellType[merged_df$cellType == "B_cells_ratio"] = "All B cells"
merged_df$cellType[merged_df$cellType == "switched_ratio"] = "Class switched"
merged_df$cellType[merged_df$cellType == "unswitched_ratio"] = "Non-class switched"
merged_df$cellType[merged_df$cellType == "NaiveB_ratio"] = "Naive"
merged_df$cellType[merged_df$cellType == "MemoryB_ratio"] = "Memory"
merged_df$cellType[merged_df$cellType == "PlasmaB_ratio"] = "Plasma"

# Create a boxplot with points colored by dataset and cellType ordered by median AUC
fig_width = 4
fig_height = 3.5
fontSize = 1.2
pdf_file <- paste0(result_fig_dir,paste0("B_and_subgroups_AUC.pdf"))
pdf(pdf_file, onefile=FALSE, width = fig_width,height=fig_height)
ggplot(merged_df, aes(x = reorder(cellType, -median_AUC), y = AUC)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(aes(color = dataset), width = 0.2, size = 2) +  
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        axis.text.y = element_text(color="black"),
        axis.text.x = element_text(color="black"),
        panel.border = element_blank(), 
        ) +
  scale_color_manual(values = custom_colors) +
  ylim(0, 1)+
  labs(title = "", x = "", y = "AUC") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()

```

