---
title: "High B cell abundance correlates with an anti-tumor microenvironment (Cohort10)"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
require(ggplot2)
ggplot2::theme_set(new = theme_bw(base_size = 12)) # sets the theme and axis font size
```


# Aims:

- Validate the relationship between B cell abundance and identified genes and signatures in an in-house cohort (Cohort10)


# load required package
```{r}

library(survival)
library(survminer)
library(ggplot2)
library(RColorBrewer)
library(glmnet)
library(rlang)
library(pROC)
library(verification)
library(tibble)
library(gridExtra)
library(forestplot)

library(GSVA)

library(org.Hs.eg.db)
library(readxl)
library(openxlsx)

library(purrr)
library(readr)

```

# common functions and constants
```{r}

process_file <- function(file_path) {
  sample_name <- gsub(".*result_(.*)\\.tsv", "\\1", basename(file_path))
  df <- read_tsv(file_path, col_names = TRUE)
  df <- df[, 1:2]  
  colnames(df) <- c("Cell_Type", sample_name)
  return(df)
}

```

# load in-house 1 data (Cohort10)
```{r}

Morris_gene_expr_df = read.csv(file = "../02.Input/MSK/Morris.HNIO.Sample.hg19KnownGene.tpm.csv", header = T, row.names = 1)

gene_symbols = mapIds(x = org.Hs.eg.db,
                                  keys=rownames(Morris_gene_expr_df),
                                  keytype="ENTREZID",
                                  column="SYMBOL",
                                  multiVals="first")

Morris_gene_expr_df$geneSymbol = gene_symbols
Morris_gene_expr_df$geneSymbol[is.na(Morris_gene_expr_df$geneSymbol)] = rownames(Morris_gene_expr_df)[is.na(Morris_gene_expr_df$geneSymbol)]

Morris_gene_expr_df = Morris_gene_expr_df[!duplicated(Morris_gene_expr_df$geneSymbol),]
rownames(Morris_gene_expr_df) = Morris_gene_expr_df$geneSymbol
Morris_gene_expr_df$geneSymbol = NULL
Morris_Kassandra_cellRatio_df = read.csv("../02.Input/MSK/Morris_Kassandra_deconvolution_result.csv", row.names = 1)
Morris_Kassandra_cellRatio_df$ID = Morris_Kassandra_cellRatio_df$Sample

############### clincal_info
clincal_info_df = read.xlsx("../02.Input/MSK/HN_preIO_RNAseq_Clinical data for nci.xlsx", sheet = "HN_Illumina.Sample_Info")

Morris_all_info = merge(Morris_Kassandra_cellRatio_df, clincal_info_df, by = "Sample")
Morris_all_info$ICBR = 0
Morris_all_info$ICBR[Morris_all_info$BEST_RESPONSE %in% c("CR", "PR")] = 1


```

# Figure S15F. Cell abundance correlation heatmap (Note that MCPcounter was used here due to its ability to estimate abundance of dendritic cells)
```{r}

tumor_ICF_df = Morris_all_info 
colnames(tumor_ICF_df)
tumor_ICF_df = tumor_ICF_df[c(3,6:12)]
cat(colnames(tumor_ICF_df), sep = '", "')
colnames(tumor_ICF_df) = c("T cells", "B cells", "NK cells", "Macrophages/Monocytes", "Dendritic cells", "Neutrophils", "Endothelial cells", "Fibroblasts")
cell_types_all = colnames(tumor_ICF_df) 

write.csv(tumor_ICF_df, file = "../02.Input/MSK/Morris_cell_abundance_MCPcounter.csv")

#### correlation heatmap plot
# Run: python CellAbundance_CorHeatmap.py

```


# Figure S15G. comparison of gene IFNG and CXCL9, and GMCSF cytosig score in B-high vs B-low tumors
```{r}

Morris_Bratio_df = Morris_Kassandra_cellRatio_df[c("ID", "B_cells")]
Morris_Bratio_df$B_score = 1
Morris_Bratio_df$B_score[Morris_Bratio_df$B_cells <= quantile (Morris_Bratio_df$B_cells, 0.50)] = 0

Morris_gene_expr_df_log1p <- log1p(Morris_gene_expr_df)
Morris_gene_expr_df_log1p = data.frame(t(Morris_gene_expr_df_log1p))
Morris_gene_expr_df_log1p$ID = rownames(Morris_gene_expr_df_log1p)
all_info_df = merge(Morris_gene_expr_df_log1p[c("ID", "IFNG", "CXCL9")], Morris_Bratio_df, by = "ID")

test_gene_vec = c("IFNG", "CXCL9")
for (test_gene in test_gene_vec) {
  all_info <- all_info_df[c("B_score", test_gene)]
  colnames(all_info) <- c("B_score", "test_gene_value")
  
  pdf(file = paste0(result_fig_dir, "Box_Morris_", test_gene, "_Bhigh_vs_Blow.pdf"), height = 2, width = 2*0.8)
  
  p <- ggplot(data = all_info) +
    geom_boxplot(aes(x = factor(B_score, levels = c(0, 1)), y = test_gene_value, fill = factor(B_score, levels = c(0, 1))), outlier.shape = NA) +
    geom_point(aes(x = B_score + 1 + runif(length(test_gene_value), min = -0.2, max = 0.2), y = test_gene_value)) +
    scale_x_discrete(labels = c("0" = "Blow", "1" = "Bhigh")) + 
    labs(x = "", y = paste0(test_gene, "")) +
    
    theme(panel.grid = element_blank(), 
          panel.background = element_rect(fill = "white"),
          axis.line.y = element_line(color = "black"),
          axis.line.x = element_line(color = "black"),
          axis.ticks.y = element_line(color = "black"),
          axis.ticks.x = element_line(color = "black"),
          axis.text.y = element_text(color = "black"),
          axis.text.x = element_text(color = "black"),
          panel.border = element_blank(), 
          legend.position = "none")
  print(p)
  dev.off()
  # Perform the Wilcoxon test and print the result
  print(paste("**************************", test_gene, "**************************"))
  print(wilcox.test(all_info$test_gene_value[all_info$B_score == 1], all_info$test_gene_value[all_info$B_score == 0])) 
}



#### load cytosig result
cytosig_result_df = read.csv(file = "../02.Input/MSK/Cytosig_heatmap.csv")
all_info_df = merge(cytosig_result_df, Morris_Bratio_df, by = "ID")

test_cytosig_vec = c( "GMCSF")
for (test_gene in test_cytosig_vec) {
  all_info <- all_info_df[c("B_score", test_gene)]
  colnames(all_info) <- c("B_score", "test_gene_value")
  
  pdf(file = paste0(result_fig_dir, "Box_Morris_", test_gene, "_Bhigh_vs_Blow.pdf"), height = 2, width = 2*0.8)
  
  p <- ggplot(data = all_info) +
    geom_boxplot(aes(x = factor(B_score, levels = c(0, 1)), y = test_gene_value, fill = factor(B_score, levels = c(0, 1))), outlier.shape = NA) +
    geom_point(aes(x = B_score + 1 + runif(length(test_gene_value), min = -0.2, max = 0.2), y = test_gene_value)) +
    scale_x_discrete(labels = c("0" = "Blow", "1" = "Bhigh")) + 
    labs(x = "", y = paste0(test_gene, "")) +
    
    theme(panel.grid = element_blank(), 
          panel.background = element_rect(fill = "white"),
          axis.line.y = element_line(color = "black"),
          axis.line.x = element_line(color = "black"),
          axis.ticks.y = element_line(color = "black"),
          axis.ticks.x = element_line(color = "black"),
          axis.text.y = element_text(color = "black"),
          axis.text.x = element_text(color = "black"),
          panel.border = element_blank(), 
          legend.position = "none")
  print(p)
  dev.off()
  # Perform the Wilcoxon test and print the result
  print(paste("**************************", test_gene, "**************************"))
  print(wilcox.test(all_info$test_gene_value[all_info$B_score == 1], all_info$test_gene_value[all_info$B_score == 0])) 
}

```


# Figure S15G. comparison of MHCII, cross-presentation, and Hypoxia ssgsea scores in B-high vs B-low tumors
```{r}

# signatures
signature_list <- list(
  ER_phago_vacuolar = c("SEC61A1","CALR", "RAB7A", "LAMP1", "NPC1", "CYBB","CTSS","CTSB","CTSL"),
  Hypoxia_Buffa = c("VEGFA", "SLC2A1", "PGAM1", "ENO1", "LDHA", "TPI1", "P4HA1", "MRPS17", "CDKN3", "ADM", "NDRG1", "TUBB6", "ALDOA", "MIF", "ACOT7", "MCTS1", "PSRC1", "PSMA7", "ANLN", "TUBA1B", "MAD2L2", "GPI", "TUBA1C", "MAP7D1", "DDIT4", "BNIP3", "C20orf20", "HIG2", "GAPDH", "MRPL13", "CHCHD2", "YKT6", "NP", "CORO1C", "SEC61G", "ANKRD37", "ESRP1", "PFKP", "SHCBP1", "CTSL2", "KIF20A", "SLC25A32", "UTP11L", "SLC16A1", "MRPL15", "KIF4A", "LRRC42", "PGK1", "HK2", "AK3L1", "CA9"),
  MHCII_genes = c("HLA-DRA", "HLA-DRB1", "HLA-DMA", "HLA-DPA1", "HLA-DPB1", "HLA-DMB", "HLA-DQB1", "HLA-DQA1", "CIITA")
)

###### GSVA signatures
pseudobulk_mat = as.matrix(Morris_gene_expr_df)

ssgsea_scores <- gsva(pseudobulk_mat, signature_list, method = "ssgsea")
rownames(ssgsea_scores) = paste0(rownames(ssgsea_scores), "_ssgsea")

ssgsea_scores_df = data.frame(t(ssgsea_scores))
ssgsea_scores_df$ID = rownames(ssgsea_scores_df)

Morris_Bratio_df = Morris_Kassandra_cellRatio_df[c("ID", "B_cells")]
Morris_Bratio_df$B_score = 1
Morris_Bratio_df$B_score[Morris_Bratio_df$B_cells <= quantile (Morris_Bratio_df$B_cells, 0.50)] = 0

all_info_df = merge(ssgsea_scores_df, Morris_Bratio_df, by = "ID")
cat(colnames(all_info_df), sep = '", "')
colnames(all_info_df) = c("ID", "Cross-presentation\n score", "Hypoxia score", "MHCII score", "B_cells", "B_score")

test_signature_vec = c("MHCII score", "Cross-presentation\n score", "Hypoxia score")
for (test_gene in test_signature_vec) {
  all_info <- all_info_df[c("B_score", test_gene)]
  colnames(all_info) <- c("B_score", "test_gene_value")
  
  pdf(file = paste0(result_fig_dir, "Box_Morris_", test_gene, "_Bhigh_vs_Blow.pdf"), height = 2, width = 2*0.8)
  p <- ggplot(data = all_info) +
    geom_boxplot(aes(x = factor(B_score, levels = c(0, 1)), y = test_gene_value, fill = factor(B_score, levels = c(0, 1))), outlier.shape = NA) +
    geom_point(aes(x = B_score + 1 + runif(length(test_gene_value), min = -0.2, max = 0.2), y = test_gene_value)) +
    scale_x_discrete(labels = c("0" = "Blow", "1" = "Bhigh")) + 
    labs(x = "", y = paste0(test_gene, "")) +
    ylim(-0.15, 1) +  
    theme(panel.grid = element_blank(), 
          panel.background = element_rect(fill = "white"),
          axis.line.y = element_line(color = "black"),
          axis.line.x = element_line(color = "black"),
          axis.ticks.y = element_line(color = "black"),
          axis.ticks.x = element_line(color = "black"),
          axis.text.y = element_text(color = "black"),
          axis.text.x = element_text(color = "black"),
          panel.border = element_blank(), 
          legend.position = "none")
  print(p)
  dev.off()
  # Perform the Wilcoxon test and print the result
  print(wilcox.test(all_info$test_gene_value[all_info$B_score == 1], all_info$test_gene_value[all_info$B_score == 0])) 
}

```



